---
title: "p8105_hw2_zf2384"
author: "Zichen Fan"
date: "2025-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse)
```

## Problem 1
Clean pols-month
```{r clean_pols}
pols_df = read_csv("./data/pols-month.csv") |>
  # separate mon into year, month, day
  separate(mon, into = c("year", "month", "day"), sep = "-") |>
  mutate(
    year = as.integer(year),
    month = as.integer(month),
    # replace month number with month name
    month = month.name[month],
    # create president variable
    president = case_when(
      prez_gop == 1 ~ "gop",
      prez_dem == 1 ~ "dem"
    )
  ) |>
  # drop day, prez_gop, prez_dem
  select(-day, -prez_gop, -prez_dem)

head(pols_df)

```
 
Clean snp
```{r clean_snp}
snp_df = read_csv("./data/snp.csv") |>
  separate(date, into = c("month", "day", "year"), sep = "/") |>
  mutate(
    year = as.integer(year),
    # adjust year if given in 2-digit format
    year = ifelse(year > 50, 1900 + year, 2000 + year),
    month = as.integer(month),
    month = month.name[month]
  ) |>
  select(year, month, close) |>
  arrange(year, month)

head(snp_df)

```

clean unemploy
```{r clean_unemploy}
unemp_df = read_csv("./data/unemployment.csv") |>
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "unemp_rate"
  ) |>
  mutate(
    month = match(month, month.abb),  # convert "Jan" → 1, etc
    month = month.name[month]
  ) |>
  rename(year = Year)
head(unemp_df)

```

Merge all data sets
```{r merge}
merged_df = pols_df |>
  left_join(snp_df, by = c("year", "month")) |>
  left_join(unemp_df, by = c("year", "month"))

merged_df

```

The pols-month data set (822 observations, 1947–2015) records counts of governors, senators, and representatives by party, plus the president’s party. The snp data set (787 observations, 1950s–2050) provides monthly closing values of the S&P stock index. The unemployment data set (816 observations, 1948–2015) contains monthly unemployment rates. After cleaning and merging, the combined data set has 822 monthly observations spanning 1947–2015, with key variables including party composition (gov_gop, sen_dem, rep_dem), president’s party, stock index values (close), and unemployment rates (unemp_rate).

## Problem 2

clean mr trash
```{r mr_trash}
library(readxl)

q2_path <- "data/202509 Trash Wheel Collection Data.xlsx"

sheet_mr_trash <- "Mr. Trash Wheel"

# Read the specified sheet, cleaning
mr_trash = read_excel(
  path = q2_path,
  sheet = sheet_mr_trash,)|>
  janitor::clean_names()|>
  select(dumpster:homes_powered)|>
   mutate(
    sports_balls = as.integer(round(sports_balls)))|>
      mutate(wheel_name = "mr_trash")|>
  mutate(year = as.numeric(year))
    
                    
```

clean professor trash and gwynnda

```{r professor_trash}
sheet_prof_trash <- "Professor Trash Wheel"

sheet_gwynns <- "Gwynns Falls Trash Wheel"

# Read the professor trash and gwynnda sheet, cleaning
prof_trash = read_excel(
  path = q2_path,
  sheet = sheet_prof_trash)|>
  janitor::clean_names() |>
   mutate(wheel_name = "prof_trash")

gwynns_trash = read_excel(
  path = q2_path,
  sheet = sheet_gwynns)|>
  janitor::clean_names() |>
   mutate(wheel_name = "gwynns_trash")

```

merge the sheets
```{r}
all_trash_data = bind_rows(mr_trash , prof_trash , gwynns_trash) 

total_obs <- nrow(all_trash_data)

# calculate total weight of professor trash wheel
prof_total_weight = all_trash_data |>
  filter(wheel_name == "prof_trash") |>
  summarise(total_weight = sum(weight_tons, na.rm = TRUE)) |>
  pull(total_weight)

# total number of cigarette butts collected by Gwynnda in June of 2022
gwynnda_cigs_june22 = all_trash_data |>
  filter(
    wheel_name == "gwynns_trash",
    year == 2022,
    month == "June"
  ) |>
  summarise(total_cigs = sum(cigarette_butts, na.rm = TRUE)) %>%
  pull(total_cigs)

```

  The combined data set contains observations from three different trash wheels in Baltimore, totaling 1191 collections. Key variables in this neat data set include wheel_name,which identifies the specific wheel, and a series of quantitative metrics such as weight_tons (tons of trash), plastic_bottles (number of plastic bottles), polystyrene (number of Styrofoam pieces), and cigarette_butts (number of cigarette butts). 

  Based on available data, the total weight of trash collected by the "Professor Trash Wheel" during its operation is approximately 564.52 tons. For a more specific question, the total number of cigarette butts collected by the Gwynns Falls trash wheel (Gwynnda) during June 2022 is 18120. 
  
## Problem 3









